---
title: "ECD WGS feature data analysis"
author:
  - "hieunguyen@genesolutions.vn"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen

---

```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.width=14, fig.height=10}
#####----------------------------------------------------------------------#####
#
# trnguyen@ukaachen.de
#
#####----------------------------------------------------------------------#####
gc()
rm(list = ls())

path.to.project.src <- "/home/hieunguyen/CRC1382/src/ecd_wgs_features"
source(file.path(path.to.project.src, "import_libraries.R"))
source(file.path(path.to.project.src, "helper_functions.R"))
library(limma)
library(umap)
library(ggpubr)
maindir <- "/home/hieunguyen/CRC1382/data"
outdir <- "/home/hieunguyen/CRC1382/data/outdir"

PROJECT <- "WGS_features"
data.version <- "20240725"
output.version <- "20240730"
path.to.main.input <- file.path(maindir, PROJECT, data.version)
path.to.main.output <- file.path(outdir, PROJECT, data.version, output.version)

path.to.01.output <- file.path(path.to.main.output, "01_output")
dir.create(path.to.01.output, showWarnings = FALSE, recursive = TRUE)

meta.data <- readxl::read_excel(Sys.glob(file.path(path.to.main.input, "metadata", "*WGS*.xlsx"))[[1]])
sample.list <- list()
for (input.label in unique(meta.data$Label)){
  sample.list[[input.label]] <- subset(meta.data, meta.data$Label == input.label)$SampleID
}
all.features <- c("GWfeature_EM",
                  "GWfeature_flen", 
                  "GWfeature_Nucleosome",
                  "GWfeature_CNA",
                  "GWfeature_flen_ratio")

unseen.runs <- c("R5434", "R5451", "R5470", "R5484")
discovery.runs <- setdiff(unique(meta.data$Run), unseen.runs)

meta.data <- meta.data %>% rowwise() %>%
  mutate(group = ifelse(Run %in% unseen.runs, "unseen", "discovery")) %>%
  mutate(group.Label = ifelse(Label == "Control", "Control", "Cancer"))

if (file.exists(file.path(path.to.01.output, "features.rds")) == FALSE){
  print("collecting features ...")
  feature.mat <- list()
  umap.mat <- list()
  
  for (feature.name in all.features){
    all.files <- Sys.glob(file.path(path.to.main.input, "feature", sprintf("*/*%s.csv", feature.name)))
    names(all.files) <- to_vec(for (item in all.files) str_split(str_split(basename(item), "_")[[1]][[1]], "-")[[1]][[2]])
    
    all.motifs <- to_vec(for (i in c("A", "T", "G", "C")) for (j in c("A", "T", "G", "C")) for (k in c("A", "T", "G", "C")) for (l in c("A", "T", "G", "C"))  sprintf("%s%s%s%s", i, j, k, l))
    
    all.files <- all.files[meta.data$SampleID]
    
    if (feature.name == "GWfeature_EM"){
      maindf <- data.frame(feat = all.motifs)
    } else if (feature.name == "GWfeature_flen"){
      maindf <- data.frame(feat = seq(1, 301))  
    } else if (feature.name == "GWfeature_Nucleosome"){
      maindf <- data.frame(feat = seq(-300, 300))
    } else if (feature.name == "GWfeature_CNA"){
      maindf <- data.frame(feat = read.csv(all.files[[1]])[, c("X")])
    } else if (feature.name == "GWfeature_flen_ratio"){
      maindf <- data.frame(feat = read.csv(all.files[[1]])[["region"]])
    }
    
    for (i in seq(1, length(all.files))){
      if (feature.name == "GWfeature_EM"){
        tmpdf <- read.csv(all.files[[i]])[, c("motif", "freq")]
        colnames(tmpdf) <- c("feat", names(all.files)[[i]])
        maindf <- merge(maindf, tmpdf, by.x = "feat", by.y = "feat")
      } else if (feature.name == "GWfeature_flen" ){
        tmpdf <- read.csv(all.files[[i]])[, c("X", "freq")]
        colnames(tmpdf) <- c("feat", names(all.files)[[i]])
        maindf <- merge(maindf, tmpdf, by.x = "feat", by.y = "feat")    
      } else if (feature.name == "GWfeature_Nucleosome"){
        tmpdf <- read.csv(all.files[[i]])
        tmpdf <- tmpdf %>% column_to_rownames("Sample") %>% t() %>% as.data.frame() %>% rownames_to_column("size")#
        tmpdf$size <- seq(-300, 300)
        colnames(tmpdf) <- c("feat", names(all.files)[[i]])
        maindf <- merge(maindf, tmpdf, by.x = "feat", by.y = "feat")    
      } else if (feature.name == "GWfeature_CNA"){
        tmpdf <- read.csv(all.files[[i]])
        colnames(tmpdf) <- c("feat", names(all.files)[[i]])
        maindf <- merge(maindf, tmpdf, by.x = "feat", by.y = "feat")    
      } else if (feature.name == "GWfeature_flen_ratio"){
        tmpdf <- read.csv(all.files[[i]])[, c("region", "ratio_short_long")]
        colnames(tmpdf) <- c("feat", names(all.files)[[i]])
        maindf <- merge(maindf, tmpdf, by.x = "feat", by.y = "feat")    
      }
    }
    maindf <- maindf %>% column_to_rownames("feat")
    if (feature.name %in% c("GWfeature_CNA", "GWfeature_flen_ratio")){
      maindf <- drop_na(maindf)
      maindf <- maindf[!is.infinite(rowSums(maindf)),]
    }
    
    umap.res <- umap(as.matrix(maindf) %>% t())
    umapdf <- data.frame(umap.res$layout) %>% as.data.frame()
    colnames(umapdf) <- c("UMAP1", "UMAP2")
    umapdf <- umapdf %>% rownames_to_column("SampleID")
    umapdf <- merge(umapdf, meta.data, by.x = "SampleID", by.y = "SampleID")
    
    feature.mat[[feature.name]] <- maindf
    umap.mat[[feature.name]] <- umapdf
  }
  saveRDS(umap.mat, file.path(path.to.01.output, "umap_res.rds"))
  saveRDS(feature.mat, file.path(path.to.01.output, "features.rds"))
} else {
  print("collected features exists! reading in ...")
  feature.mat <- readRDS(file.path(path.to.01.output, "features.rds"))
  umap.mat <- readRDS(file.path(path.to.01.output, "umap_res.rds"))
}

count.label.in.run <- table(meta.data$Label, meta.data$Run) %>% as.data.frame()
count.cancer.control.in.run <- table(meta.data$group.Label, meta.data$Run) %>% as.data.frame()
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.width=14, fig.height=10}
##### save features mat to csv file
dir.create(file.path(path.to.01.output, "csv"), showWarnings = FALSE, recursive = TRUE)

for (f in names(feature.mat)){
  tmpdf <- feature.mat[[f]]
  write.csv(tmpdf, file.path(path.to.01.output, "csv", sprintf("%s.csv", f)))
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
count.label.in.run %>% ggplot(aes(x = Var2, y = Freq, fill = Var1)) + geom_bar(stat = "identity") + facet_wrap(~Var1) + theme_pubr(x.text.angle = 90, base_size = 20)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
count.cancer.control.in.run %>% ggplot(aes(x = Var2, y = Freq, fill = Var1)) + geom_bar(stat = "identity") + facet_wrap(~Var1) + theme_pubr(x.text.angle = 90, base_size = 20)
```

# UMAP
## All samples, group by run {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in names(umap.mat)){
  cat(sprintf("### Feature %s \n", feat))
  p <- ggplot(data = umap.mat[[feat]], aes(x = UMAP1, y = UMAP2, color = Run)) + geom_point(size = 3) + theme_pubr()
  print(p)
  cat("\n \n")
}
```

## All samples, group by label {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in names(umap.mat)){
  cat(sprintf("### Feature %s \n", feat))
  p <- ggplot(data = umap.mat[[feat]], aes(x = UMAP1, y = UMAP2, color = Label)) + geom_point(size = 3) + theme_pubr()
  print(p)
  cat("\n \n")
}
```

## All samples, group by discovery and unseen {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in names(umap.mat)){
  cat(sprintf("### Feature %s \n", feat))
  p <- ggplot(data = umap.mat[[feat]], aes(x = UMAP1, y = UMAP2, color = group)) + geom_point(size = 3) + theme_pubr()
  print(p)
  cat("\n \n")
}
```

## Control samples only {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in all.features){
  cat(sprintf("### Feature %s \n", feat))
  tmpdf <- feature.mat[[feat]][, sample.list$Control]
  umap.res <- umap(as.matrix(tmpdf) %>% t())
  umapdf <- data.frame(umap.res$layout) %>% as.data.frame()
  colnames(umapdf) <- c("UMAP1", "UMAP2")
  umapdf <- umapdf %>% rownames_to_column("SampleID")
  umapdf <- merge(umapdf, meta.data, by.x = "SampleID", by.y = "SampleID")
  p <- ggplot(data = umapdf, aes(x = UMAP1, y = UMAP2, color = group)) + geom_point(size = 3) + theme_pubr()
  print(p)  
  cat("\n \n")
}
```

## Control samples only, grouped by Run {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in all.features){
  cat(sprintf("### Feature %s \n", feat))
  tmpdf <- feature.mat[[feat]][, sample.list$Control]
  umap.res <- umap(as.matrix(tmpdf) %>% t())
  umapdf <- data.frame(umap.res$layout) %>% as.data.frame()
  colnames(umapdf) <- c("UMAP1", "UMAP2")
  umapdf <- umapdf %>% rownames_to_column("SampleID")
  umapdf <- merge(umapdf, meta.data, by.x = "SampleID", by.y = "SampleID")
  p <- ggplot(data = umapdf, aes(x = UMAP1, y = UMAP2, color = Run)) + geom_point(size = 3) + theme_pubr()
  print(p)  
  cat("\n \n")
}
```

## Summary: the number of samples in each set
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
table(meta.data$group, meta.data$Label)
```

## Breast samples only {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in all.features){
  cat(sprintf("### Feature %s \n", feat))
  tmpdf <- feature.mat[[feat]][, sample.list$Breast]
  umap.res <- umap(as.matrix(tmpdf) %>% t())
  umapdf <- data.frame(umap.res$layout) %>% as.data.frame()
  colnames(umapdf) <- c("UMAP1", "UMAP2")
  umapdf <- umapdf %>% rownames_to_column("SampleID")
  umapdf <- merge(umapdf, meta.data, by.x = "SampleID", by.y = "SampleID")
  p <- ggplot(data = umapdf, aes(x = UMAP1, y = UMAP2, color = group)) + geom_point(size = 3) + theme_pubr()
  print(p)  
  cat("\n \n")
}
```

## CRC samples only {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (feat in all.features){
  cat(sprintf("### Feature %s \n", feat))
  tmpdf <- feature.mat[[feat]][, sample.list$CRC]
  umap.res <- umap(as.matrix(tmpdf) %>% t())
  umapdf <- data.frame(umap.res$layout) %>% as.data.frame()
  colnames(umapdf) <- c("UMAP1", "UMAP2")
  umapdf <- umapdf %>% rownames_to_column("SampleID")
  umapdf <- merge(umapdf, meta.data, by.x = "SampleID", by.y = "SampleID")
  p <- ggplot(data = umapdf, aes(x = UMAP1, y = UMAP2, color = group)) + geom_point(size = 3) + theme_pubr()
  print(p)  
  cat("\n \n")
}
```

# Run statistics tests
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
count.labels <- table(meta.data$Label)

selected.cancer.classes <- setdiff(names(count.labels[count.labels >= 20]), c("Control"))

diff.test <- hash()
for (feature.name in all.features){
  diff.test[[feature.name]] <- list()
  for (input.cancer.label in selected.cancer.classes){
    input.mat <- feature.mat[[feature.name]]
    group1 <- sample.list[[input.cancer.label]]
    group2 <- sample.list[["Control"]]
    
    input.metadata <- data.frame(sample = c(group1, group2),
                                 label = c(
                                   to_vec(for(item in seq(1, length(group1))) 1),
                                   to_vec(for(item in seq(1, length(group2))) 0)
                                   ))
    exprsdf <- input.mat[, input.metadata$sample]
    design.mat <- model.matrix(~label, data = input.metadata)
    fit <- lmFit(exprsdf, design = design.mat)
    fit <- eBayes(fit)
    stats_df <- topTable(fit, number = nrow(exprsdf), coef = "label") %>%
      tibble::rownames_to_column("region_name") %>%
      rowwise() %>%
      mutate(abs_logFC = abs(logFC))
    stats_df <- subset(stats_df, is.na(stats_df$adj.P.Val) ==  FALSE) %>% subset(adj.P.Val <= 0.05) %>% arrange(desc(abs_logFC)) 
    
    diff.test[[feature.name]][[input.cancer.label]] <- head(stats_df, round(0.2 * nrow(feature.mat[[feature.name]]))) %>% pull("region_name")
  } 
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
library(UpSetR)
library(grid)
for (feature.name in all.features){
  listInput <- diff.test[[feature.name]]
  print(upset(fromList(listInput), order.by = c("freq"), decreasing = c(TRUE,TRUE), nsets = length(listInput), point.size = 5, set_size.numbers_size = 12, set_size.show = 12, text.scale = 3))
  grid.text(feature.name, x = 0.65, y=0.95, gp=gpar(fontsize=20))
}
```

# How to diagnose feature drift or batch effect?
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}

control.metadata <- subset(meta.data, meta.data$Label == "Control") %>% arrange(desc(Run))

all.runs <- unique(control.metadata$Run) %>% sort()

ref.run <- all.runs[[1]]
```

## Median shift
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
library(gganimate)
# 
# for (feature.name in all.features){
#   maindf <- feature.mat[[feature.name]][, sample.list$Control] 
#   plotdf <- data.frame(feat = row.names(maindf))
#   
#   for (run.id in unique(control.metadata$Run)){
#     tmpdf <- maindf[, subset(control.metadata, control.metadata$Run == run.id)$SampleID]
#     tmpdf$median <- rowMedians(as.matrix(tmpdf)) 
#     tmpdf <- tmpdf %>% as.data.frame() %>% rownames_to_column("feat") %>% subset(select = c(feat, median))
#     colnames(tmpdf) <- c("feat", run.id)
#     plotdf <- merge(plotdf, tmpdf, by.x = "feat", by.y = "feat")
#   }
#   
#   long.plotdf <- data.frame()
#   for (run.id in setdiff(all.runs, ref.run)){
#     tmpdf <- plotdf[, c(ref.run, run.id)]
#     tmpdf$Run <- as.numeric(str_replace(run.id, "R", ""))
#     colnames(tmpdf) <- c("x", "y", "Run")
#     long.plotdf <- rbind(long.plotdf, tmpdf)
#   }
#   
#   p.anim <- ggplot(data = long.plotdf, aes(x = x, y = y)) + geom_point() + transition_time(Run) + ggtitle(feature.name)
#   anim_save(filename = sprintf("median_shift_%s.gif", feature.name), animation = p.anim, path = path.to.01.output)
# }
```

## Std shift
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
# for (feature.name in all.features){
#   maindf <- feature.mat[[feature.name]][, sample.list$Control] 
#   plotdf <- data.frame(feat = row.names(maindf))
#   
#   for (run.id in unique(control.metadata$Run)){
#     tmpdf <- maindf[, subset(control.metadata, control.metadata$Run == run.id)$SampleID]
#     tmpdf$std <- rowSds(as.matrix(tmpdf)) 
#     tmpdf <- tmpdf %>% as.data.frame() %>% rownames_to_column("feat") %>% subset(select = c(feat, std))
#     colnames(tmpdf) <- c("feat", run.id)
#     plotdf <- merge(plotdf, tmpdf, by.x = "feat", by.y = "feat")
#   }
#   
#   long.plotdf <- data.frame()
#   for (run.id in setdiff(all.runs, ref.run)){
#     tmpdf <- plotdf[, c(ref.run, run.id)]
#     tmpdf$Run <- as.numeric(str_replace(run.id, "R", ""))
#     colnames(tmpdf) <- c("x", "y", "Run")
#     long.plotdf <- rbind(long.plotdf, tmpdf)
#   }
#   
#   p.anim <- ggplot(data = long.plotdf, aes(x = x, y = y)) + geom_point() + transition_time(Run) + ggtitle(feature.name)
#   anim_save(filename = sprintf("std_shift_%s.gif", feature.name), animation = p.anim, path = path.to.01.output)
# }
```

## UMAP shift
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
```

## Stats test
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=10}
maindf <- feature.mat[[feature.name]][, sample.list$Control] 
discoverydf <- maindf[, subset(control.metadata, control.metadata$Run %in% discovery.runs)$SampleID]
unseendf <- maindf[, subset(control.metadata, control.metadata$Run %in% unseen.runs)$SampleID]

f <- "1:100000001-101000000"
x <- discoverydf[f,] %>% unlist()
y <- unseendf[f, ] %>% unlist()

hist(x , freq = FALSE, main = "Histogram and density", color = "red")
hist(y , freq = FALSE, main = "Histogram and density", color = "blue")

# Calculate density
dx <- density(x)
dy <- density(y)

# Add density
lines(dx, lwd = 2, col = "red")
lines(dy, lwd = 2, col = "blue")

# Plot the density without histogram
plot(dx, lwd = 2, col = "red",
     main = "Density")
plot(dy, lwd = 2, col = "blue",
     main = "Density")
```

## VGAM test

